A Computational Account of Default Mode Function\\ by Reinforcement Learning Theory

### 2017-03-26 paper/paper.tex
more notes

\subsection{The hippocampus: Memory, space, and experience replay}
The HC is well known to be involved in memory and spatial navigation in animals and humans.
Its highly recursive anatomical architecture
may be specifically designed to allow reconstructing
entire episodes of experience from memory fragments.
%
While the HC in the medial temporal lobe system
is traditionally believed to allow remembering the past,
there is now increasing evidence for a role
in constructing mental models in general
\citep{maguire2016, schacter2007remembering}.
Indeed,
hippocampal damage is
not only associated with an impairment in reexperiencing the past (i.e., amnesia),
but also thinking about one’s own future and
imagining new fictitious experiences more broadly \citep{hassabis2007patients}.
Mental scenes created by hippocampus-lesioned patients exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal hippocampus revealed
some constantly active neuronal ensembles whose firing coincided with
distinct locations in space while the animal navigated through its surroundings.
London taxi drivers, individuals with high performance in spatial navigation,
were shown to exhibit increased grey matter volume in the
posterior hippocampus \citep{maguire2000navigation}.
But encoding and generative reconstruction in the hippocampus extends
beyond mere spatial knowledge of the environment.
Based on large-scale recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, that were elicited in experimental task conditions,
have been shown to be reenacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space}.
Moreover, spike sequences measured in hippocampal place cells of rats
featured reoccurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Such hippocampal ensemble burts during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the nodes of the DMN.
These mechanisms of the hippocampus probably make important contributions to the
recollection of autobiographical memory episodes and other
reexperienced or newly generated mental scenarios
\citep{hassabis2007patients}.
%
The HC thus orchestrates elements of experienced environmental aspects for
consolidations based on reenactment and for integration into
rich mental scene construction. In this way, the HC may even influence
ongoing perception in the current environment
\citep{maguire2016}.
% (TODO: relate to self-play in deepmind Go).
% (TODO: Kai had an interesting idea: the deja-vu effect
% could be when you already accidentally perturbed past
% such that it really occures in reality later.)

@@

\section{Relation to other statistical models (Elvis)}
\subsection{Friston's Free-energy principle (refer to \ref{label:friston})}
\textbf{XXX: add tons of more examples and connnections}
\textbf{XXX: integrate aspects from section "ThE MARkOv DECISION PROBLEM" in Dayan and Daw}
\textbf{XXX: integrate aspects from WakeNsleep by Hinton}
In summary:
\begin{itemize}
  \item Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  \item A common point between the Free Energy principle and our framework
  is placing the minimization of suprise at the core of brain function.
  In RL, surprise minimization reduces to accurate prediction of
  rewarding outcomes in the future.
  \item Ortega et al. 2013 \citep{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \citep{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
\item It is difficult to compute with the free energy principle
  in practice, which makes it physiologically less plausible. Auto-encoder theory presented
  \ref{sec:vae} can help to remove these barriers, by making the whole scheme trainable via
  backprop!
  \end{itemize}

@@

\section{Relation to other cognitive models}
\subsection{Predictive Coding Hypothesis}
The predictive coding and Bayesian brain framework
is a frequently evoked idea related to the default mode function
\citep{bar2007}.
According to this cognitive account,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information.
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
In fact, axonal
back projections probably outnumber by far the input projection
existing in the brain \citep{salin1995corticocortical}.
This can explain the constructive, generative nature of sensory perception
\citep{friston2010free} and
why motor action is intimately related to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is hence maintained by top-down modulation of sensorimotor
processing by a-priori information about the environment.
The generative model of how perceived sensory events arise in the
world would be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the constant influx of sensory examples.

>>>>

\subsection{The hippocampus: Memory, space, and experience replay}
The HC is well known to be involved in memory and spatial navigation in animals and humans.
Its highly recursive anatomical architecture
may be specifically designed to allow reconstructing
entire episodes of experience from memory fragments.
%
While the HC in the medial temporal lobe system
is traditionally believed to allow remembering the past,
there is now increasing evidence for a role
in constructing mental models in general
\citep{maguire2016, schacter2007remembering}.
Indeed,
hippocampal damage is
not only associated with an impairment in reexperiencing the past (i.e., amnesia),
but also thinking about one’s own future and
imagining new fictitious experiences more broadly \citep{hassabis2007patients}.
Mental scenes created by hippocampus-lesioned patients exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal hippocampus revealed
some constantly active neuronal ensembles whose firing coincided with
distinct locations in space while the animal navigated through its surroundings.
London taxi drivers, individuals with high performance in spatial navigation,
were shown to exhibit increased grey matter volume in the
posterior hippocampus \citep{maguire2000navigation}.
Indeed, place-specific neuronal populations in the hippocampus
spike one after another when an animal is choosing between alternative
paths \citep{johnson2007neural}; and
these neuronal patterns apear to also indicate upcoming behavior
\citep{pfeiffer2013hippocampal}.

But encoding and generative reconstruction in the hippocampus extends
beyond mere spatial knowledge of the environment.
Based on large-scale recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, that were elicited in experimental task conditions,
have been shown to be reenacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space}.
Moreover, spike sequences measured in hippocampal place cells of rats
featured reoccurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Such hippocampal ensemble burts during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the nodes of the DMN.
These mechanisms of the hippocampus probably make important contributions to the
recollection of autobiographical memory episodes and other
reexperienced or newly generated mental scenarios
\citep{hassabis2007patients}.
%
The HC thus orchestrates elements of experienced environmental aspects for
consolidations based on reenactment and for integration into
rich mental scene construction. In this way, the HC may even influence
ongoing perception in the current environment
\citep{maguire2016}.
% (TODO: relate to self-play in deepmind Go).
% (TODO: Kai had an interesting idea: the deja-vu effect
% could be when you already accidentally perturbed past
% such that it really occures in reality later.)

@@

\section{Relation to other statistical models (Elvis)}
\subsection{Friston's Free-energy principle (refer to \ref{label:friston})}
\textbf{XXX: add tons of more examples and connnections}
\textbf{XXX: integrate aspects from section "ThE MARkOv DECISION PROBLEM" in Dayan and Daw}
\textbf{XXX: integrate aspects from WakeNsleep by Hinton}
\textbf{XXX: integrate aspects from Sutton/Barto book chapter}
In summary:
\begin{itemize}
  \item Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  \item A common point between the Free Energy principle and our framework
  is placing the minimization of suprise at the core of brain function.
  In RL, surprise minimization reduces to accurate prediction of
  rewarding outcomes in the future.
  \item Ortega et al. 2013 \citep{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \citep{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
\item It is difficult to compute with the free energy principle
  in practice, which makes it physiologically less plausible. Auto-encoder theory presented
  \ref{sec:vae} can help to remove these barriers, by making the whole scheme trainable via
  backprop!
  \end{itemize}

@@

\section{Relation to other cognitive models}
\subsection{Predictive Coding Hypothesis}
The predictive coding framework
\citep{clark2013whatever}
is a frequently evoked idea related to the default mode function
\citep{bar2007}.
According to this cognitive account,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information.
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
In fact, axonal
back projections probably outnumber by far the input projection
existing in the brain \citep{salin1995corticocortical}.
This can explain the constructive, generative nature of sensory perception
\citep{friston2010free} and
why motor action is intimately related to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is hence maintained by top-down modulation of sensorimotor
processing by a-priori information about the environment.
The generative model of how perceived sensory events arise in the
world would be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the constant influx of sensory examples.
